\textbf{Differences in COVID-19 vaccination in the province of Ontario across Health Regions and socio-economic strata}

### 2023-06-27 manuscript_v5/01-background.qmd
v6 after SW's comments

In particular, a major weakness that has received attention during the pandemic has been related to inequalities in vaccine uptake. The rapid development of vaccines against COVID-19 initially brought the hope of a rapid end to the pandemic due to the start of vaccination campaigns in certain parts of the world toward the end of 2020 [@thelancet2021;@davis2022;@bogoch2022;@tanne2020]) but inequalities in vaccine uptake made these pharmaceutical interventions ultimately unable to replicate the success of the smallpox vaccination program, which was crucial to control this disease [@kayser2021].

This problematic is a multifaceted issue resulting from a combination of factors, among which are failed public health measures [@li2021], inequality in vaccine access between high- and low-income countries [@gerretsen2021; @tamey2022], and vaccine hesitancy [@nafilyan2021]. Furthermore, it is well established that this issue has affected in particular individuals in certain underrepresented groups (e.g., Black, Asian, or Indigenous) as well as individuals with socio-economic disadvantages [@willis2021;@skirrow2022;@stoler2021;@khubchandani2021;@guay2022;@muhajarine2021;@hussain2022].

Reasons given for this inequality have included medical mistrust due to systemic medical racism [@stoler2021;@mosby2021], mistrust in vaccines [@willis2021], and the influence of conspiracy theories [@bogart2021;@mosby2021;@freeman2020].  However, it is important also to consider that vaccination uptake can be influenced by geographical (spatial) factors. In this regard, differences in COVID-19 vaccination rates have been associated with varied regional attitudes towards vaccination [@malik2020], spatial differences in vaccine access and supply, vaccination location availability, and lack of prioritization of areas where vulnerable groups reside [@bogoch2022; @nguyen2021]. Other studies have also shown heterogeneity in vaccine uptake within small governmental administrative units such as counties [@mollalo2021; @yang2022; @tiu2022; @bhuiyan2022], and that accounting for geographical differences in vaccination can help predict patterns of booster uptake [@wood2022].

>>>>

In particular, a major weakness that has received attention during the pandemic has been related to vaccination. The rapid development of vaccines against COVID-19 initially brought the hope of a rapid end to the pandemic as vaccination campaigns in certain parts of the world started toward the end of 2020[@thelancet2021;@davis2022;@bogoch2022;@tanne2020]. However, although it has been estimated that COVID-19 vaccines have been able to prevent millions of deaths worldwide [@watson2022], they were ultimately unable to bring the pandemic to a rapid end, in contrast to previous experiences (such as in the case of smallpox), where the global vaccination program was crucial to control this disease[@kayser2021].

This failure of vaccines to stop the pandemic is a complex issue resulting from a combination of factors, among which are inadequate public health measures and the emergence of new variants [@li2021], inequality in vaccine access between high- and low-income countries [@gerretsen2021;@tamey2022], vaccine hesitancy [@nafilyan2021], and differences in vaccine uptake across the population[@willis2021;@skirrow2022]. In particular, it is well established that lower vaccination uptake has been observed in individuals within certain underrepresented groups (e.g., Black, Asian, or Indigenous) as well as individuals with socio-economic disadvantages [@willis2021;@skirrow2022;@stoler2021;@khubchandani2021;@guay2022;@muhajarine2021;@hussain2022].

Reasons given for this inequality in vaccination uptake have included medical mistrust due to systemic medical racism [@stoler2021;@mosby2021], mistrust in vaccines [@willis2021], and the influence of conspiracy theories [@bogart2021;@mosby2021;@freeman2020].  However, it is important also to consider that vaccination uptake can be influenced by geographical (spatial) factors. In this regard, differences in COVID-19 vaccination rates have been associated with varied regional attitudes towards vaccination [@malik2020], spatial differences in vaccine access and supply, vaccination location availability, and lack of prioritization of areas where vulnerable groups reside [@bogoch2022; @nguyen2021]. Other studies have also shown heterogeneity in vaccine uptake within small governmental administrative units such as counties [@mollalo2021; @yang2022; @tiu2022; @bhuiyan2022], and that accounting for geographical differences in vaccination can help predict patterns of booster uptake [@wood2022].

### 2023-06-27 manuscript_v6/01-background.qmd
v6 after SW's comments

In particular, a major weakness that has received attention during the pandemic has been related to vaccination. The rapid development of vaccines against COVID-19 initially brought the hope of a rapid end to the pandemic as vaccination campaigns in certain parts of the world started toward the end of 2020[@thelancet2021;@davis2022;@bogoch2022;@tanne2020]. However, although it has been estimated that COVID-19 vaccines have been able to prevent millions of deaths worldwide [@watson2022], they were ultimately unable to bring the pandemic to a rapid end, in contrast to previous experiences (such as in the case of smallpox), where the global vaccination program was crucial to control this disease[@kayser2021].

This failure of vaccines to stop the pandemic is a complex issue resulting from a combination of factors, among which are inadequate public health measures and the emergence of new variants [@li2021], inequality in vaccine access between high- and low-income countries [@gerretsen2021;@tamey2022], vaccine hesitancy [@nafilyan2021], and differences in vaccine uptake across the population[@willis2021;@skirrow2022]. In particular, it is well established that lower vaccination uptake has been observed in individuals within certain underrepresented groups (e.g., Black, Asian, or Indigenous) as well as individuals with socio-economic disadvantages [@willis2021;@skirrow2022;@stoler2021;@khubchandani2021;@guay2022;@muhajarine2021;@hussain2022].

Reasons given for this inequality in vaccination uptake have included medical mistrust due to systemic medical racism [@stoler2021;@mosby2021], mistrust in vaccines [@willis2021], and the influence of conspiracy theories [@bogart2021;@mosby2021;@freeman2020].  However, it is important also to consider that vaccination uptake can be influenced by geographical (spatial) factors. In this regard, differences in COVID-19 vaccination rates have been associated with varied regional attitudes towards vaccination [@malik2020], spatial differences in vaccine access and supply, vaccination location availability, and lack of prioritization of areas where vulnerable groups reside [@bogoch2022; @nguyen2021]. Other studies have also shown heterogeneity in vaccine uptake within small governmental administrative units such as counties [@mollalo2021; @yang2022; @tiu2022; @bhuiyan2022], and that accounting for geographical differences in vaccination can help predict patterns of booster uptake [@wood2022].

@@

>>>>

In particular, a major weakness that has received attention during the pandemic has been related to vaccination. The rapid development of vaccines against COVID-19 initially brought the hope of a rapid end to the pandemic as vaccination campaigns in certain parts of the world started as early as the end of 2020 [@thelancet2021;@davis2022;@bogoch2022;@tanne2020]. However, although it has been estimated that COVID-19 vaccines have been able to prevent millions of deaths worldwide [@watson2022], their original goal, which was to bring the pandemic to a rapid end, was not accomplished. This contrast with previous vaccination experiences, such as in the case of smallpox, where the implementation of a global vaccination program was crucial to control this disease[@kayser2021].

The failure of vaccines to stop the pandemic is a complex issue resulting from a combination of factors, among which are inadequate public health measures and the emergence of new variants [@li2021], inequality in vaccine access between high-income and low-income countries [@gerretsen2021;@tamey2022], vaccine hesitancy [@nafilyan2021], and differences in vaccine uptake across the population[@willis2021;@skirrow2022]. In particular, it is well established that lower vaccination uptake has been observed in individuals within certain underrepresented groups (e.g., Black, Asian, or Indigenous) as well as individuals with socio-economic disadvantages [@willis2021;@skirrow2022;@stoler2021;@khubchandani2021;@guay2022;@muhajarine2021;@hussain2022].

Reasons given for this inequality in vaccination uptake have included medical mistrust due to systemic medical racism [@stoler2021;@mosby2021], mistrust in vaccines [@willis2021], and the influence of misinformation and disinformation [@bogart2021;@mosby2021;@freeman2020]. However, it is important also to consider that vaccination uptake can be influenced by geographical (spatial) factors. In this regard, differences in COVID-19 vaccination rates have been associated with varied regional attitudes towards vaccination [@malik2020], lack of vaccine access in areas where underrepresented groups reside[@bogoch2022], as well as an absence of prioritization of areas inhabited by vulnerable groups [@nguyen2021]. Other studies have also shown heterogeneity in vaccine uptake within small governmental administrative units such as counties [@mollalo2021; @yang2022; @tiu2022; @bhuiyan2022], indicating that accounting for geographical differences in vaccination can help predict patterns of booster uptake [@wood2022].

@@

On the other hand, public health is managed in the province by Public Health Ontario, which was established in late 2007 and is currently composed of 34 Public Health Units (PHUs). Importantly, during the COVID-19 pandemic, these PHUS led the distribution of the vaccine within the areas they covered [@rotenberg2021]

- add about intersection of Health Regions and PHUs. How vaccination presented a challenge as vaccine was ultimately administered in facilites managed by OHT?

https://www.mcmasterforum.org/docs/default-source/product-documents/rapid-responses/examining-the-intersections-between-ontario-health-teams-and-public-health.pdf?sfvrsn=69caa8c9_11

### 2023-06-27 manuscript_v6/03-results.qmd
v6 after SW's comments

Finally, **the place of habitation affected the odds of vaccination for certain underrepresented groups, as significantly lower odds of vaccination were identified for the interaction between Health Region and race** in the case of Black individuals in the Central Health Region (OR=0.39, CI=[0.2,0.75]), Arab/Middle Eastern individuals in the East Health Region (OR=0.41 [0.17, 0.98]), and in the Indigenous and mixed groups in the West Health Region (ORs and CIs=[0.31 [0.14, 0.7] and 0.38 [0.19, 0.76], respectively).

>>>>

Finally, the place of habitation affected the odds of vaccination for certain underrepresented groups, as significantly lower odds of vaccination were identified for the interaction between Health Region and race in the case of Black individuals in the Central Health Region (OR=0.39, CI=[0.2,0.75]), Arab/Middle Eastern individuals in the East Health Region (OR=0.41 [0.17, 0.98]), and in the Indigenous and mixed groups in the West Health Region (ORs and CIs=[0.31 [0.14, 0.7] and 0.38 [0.19, 0.76], respectively).

### 2023-06-27 manuscript_v6/04-discussion.qmd
v6 after SW's comments

In this study, we hypothesized that differences in COVID-19 vaccination uptake were present between the Health Regions between late 2021 and early 2022. Our goal was to determine which socio-demographic groups could be impacted by these disparities in order to provide decision-makers with information that could be used to develop policies focused on reducing or eliminating these differences and ensure that the Health Region model is able to fulfill its mission of improving health access for all Ontarians. **making corrections up to this point**

>>>>

In this study, we hypothesized that differences in COVID-19 vaccination uptake were present between the Health Regions between late 2021 and early 2022. Our goal was to determine which socio-demographic groups could be impacted by these disparities in order to provide decision-makers with information that could be used to develop policies focused on reducing or eliminating these differences and ensuring that the Health Region model is able to fulfill its mission of improving health access for all Ontarians.
